---
title: "Unsupervised Segmentation of Credit Card Applicants: K-Means, Hierarchical, DBSCAN, and Mean Shift Clustering"
author: "Mintay Misgano"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 3
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
```

---

## Overview

Unsupervised clustering is one of the most practical tools in an analyst's toolkit for consumer segmentation. Unlike classification, it requires no labeled outcome — it discovers latent structure in behavior patterns that can inform risk tiering, product targeting, and policy decisions.

In this analysis I apply four clustering algorithms to a credit card applicant dataset (N = 1,319) to identify behaviorally distinct applicant profiles based on income, spending share, expenditure, account tenure, and derogatory reports. The methods — K-Means, Hierarchical (Ward's D2), DBSCAN, and Mean Shift — each recover structure under different assumptions, making convergence across methods a meaningful indicator of robust segment boundaries.

---

## Setup

```{r packages}
library(tidyverse)
library(cluster)
library(dbscan)
library(meanShiftR)
library(ggplot2)
library(factoextra)
library(psych)
library(corrplot)
library(knitr)
library(scales)
```

---

## Data Preparation

```{r load-data}
df_raw <- read.csv("01_credit_card.csv", row.names = 1)

# Preview
glimpse(df_raw)
```

```{r prep-data}
# Select numeric features relevant to behavioral segmentation
# Exclude the binary outcome (card) and binary demographics (owner, selfemp)
# to ensure clustering reflects behavioral/financial patterns, not approval status
features <- c("reports","age","income","share","expenditure",
               "dependents","months","majorcards","active")

df_num <- df_raw[, features]

# Scale to zero mean, unit variance — required for distance-based methods
df_scaled <- scale(df_num)
df_scaled <- as.data.frame(df_scaled)

cat("Observations:", nrow(df_scaled), "\n")
cat("Features used:", ncol(df_scaled), "\n")
cat("Features:", paste(features, collapse=", "), "\n")
```

```{r descriptives}
df_num |>
  describe() |>
  round(2) |>
  kable(caption = "Table 1. Descriptive Statistics — Clustering Features")
```

```{r correlation-plot, fig.cap="Figure 1. Feature correlations before clustering"}
corr_mat <- cor(df_num)
corrplot::corrplot(corr_mat, method = "color", type = "lower",
                   tl.cex = 0.8, number.cex = 0.7, addCoef.col = "black",
                   col = colorRampPalette(c("#2196F3","white","#F44336"))(200))
title("Feature Correlation Matrix", cex.main = 0.95)
```

---

## Method 1: K-Means Clustering

### Determining Optimal K — Elbow Method

```{r kmeans-elbow, fig.cap="Figure 2. Elbow plot for optimal K (K-Means)"}
set.seed(100)

sse <- function(k) {
  kmeans(df_scaled, centers = k, nstart = 25, iter.max = 100)$tot.withinss
}

k_vals <- 1:12
all_sses <- map_dbl(k_vals, sse)

elbow_df <- data.frame(k = k_vals, SSE = all_sses)

ggplot(elbow_df, aes(x = k, y = SSE)) +
  geom_line(color = "#1976D2", linewidth = 0.8) +
  geom_point(size = 3, color = "#1976D2") +
  geom_vline(xintercept = 4, linetype = "dashed", color = "#F44336", alpha = 0.7) +
  annotate("text", x = 4.3, y = max(all_sses)*0.85,
           label = "K = 4 (elbow)", color = "#F44336", size = 3.5, hjust = 0) +
  scale_x_continuous(breaks = k_vals) +
  labs(title   = "Elbow Method — Total Within-Cluster SSE vs. K",
       subtitle = "Inflection point suggests K = 4 optimal segments",
       x = "Number of Clusters (K)", y = "Total Within-Cluster SSE") +
  theme_minimal(base_size = 12)
```

### K = 4 Solution

```{r kmeans-fit}
set.seed(100)
km4 <- kmeans(df_scaled, centers = 4, nstart = 25, iter.max = 100)

cat("K-Means cluster sizes:\n")
print(table(km4$cluster))
cat("\nWithin-cluster SS:", round(km4$tot.withinss, 2), "\n")
cat("Between-cluster SS:", round(km4$betweenss, 2), "\n")
cat("Ratio (between/total):", round(km4$betweenss / km4$totss, 3), "\n")
```

```{r kmeans-viz, fig.cap="Figure 3. K-Means clusters visualised on first two principal components"}
fviz_cluster(km4, data = df_scaled,
             palette = c("#1976D2","#388E3C","#F57C00","#7B1FA2"),
             geom = "point", ellipse.type = "convex",
             ggtheme = theme_minimal(base_size = 11),
             main = "K-Means (K = 4) — PCA Projection")
```

### Cluster Profiles

```{r kmeans-profile}
df_raw$km_cluster <- factor(km4$cluster)

cluster_profile <- df_raw |>
  group_by(km_cluster) |>
  summarise(
    n           = n(),
    pct_approved = mean(card == "yes") * 100,
    mean_income  = mean(income),
    mean_exp     = mean(expenditure),
    mean_share   = mean(share),
    mean_age     = mean(age),
    mean_months  = mean(months),
    mean_active  = mean(active),
    mean_reports = mean(reports),
    .groups = "drop"
  ) |>
  mutate(across(where(is.numeric), ~round(., 2)))

kable(cluster_profile,
      col.names = c("Cluster","n","% Approved","Avg Income","Avg Expenditure",
                    "Avg Share","Avg Age","Avg Tenure (mo)","Avg Active Accts","Avg Reports"),
      caption = "Table 2. K-Means Cluster Profiles — Mean Values on Key Variables")
```

---

## Method 2: Hierarchical Clustering (Ward's D2)

```{r hclust-fit, fig.cap="Figure 4. Dendrogram — Ward's D2 linkage, 4-cluster solution"}
set.seed(100)

# Use a random sample of 200 for readable dendrogram; full dataset for final assignments
set.seed(42)
sample_idx <- sample(nrow(df_scaled), 200)
df_sample  <- df_scaled[sample_idx, ]

dist_mat <- dist(df_sample, method = "euclidean")
hc_ward  <- hclust(dist_mat, method = "ward.D2")

plot(hc_ward, cex = 0.4, hang = -1,
     main = "Hierarchical Clustering — Ward's D2 Linkage (n = 200 sample)",
     xlab = "", sub = "")
rect.hclust(hc_ward, k = 4, border = c("#1976D2","#388E3C","#F57C00","#7B1FA2"))
```

```{r hclust-full}
# Assign full dataset to 4 clusters via hclust on full distance matrix
dist_full  <- dist(df_scaled, method = "euclidean")
hc_full    <- hclust(dist_full, method = "ward.D2")
hc_labels  <- cutree(hc_full, k = 4)

cat("Hierarchical cluster sizes (full N = 1319):\n")
print(table(hc_labels))
```

```{r hclust-viz, fig.cap="Figure 5. Hierarchical clusters on PCA projection"}
fviz_cluster(list(data = df_scaled, cluster = hc_labels),
             palette = c("#1976D2","#388E3C","#F57C00","#7B1FA2"),
             geom = "point", ellipse.type = "convex",
             ggtheme = theme_minimal(base_size = 11),
             main = "Hierarchical Clustering (Ward's D2, K = 4) — PCA Projection")
```

---

## Method 3: DBSCAN

DBSCAN (Density-Based Spatial Clustering of Applications with Noise) identifies clusters as dense regions separated by sparse space. Unlike K-Means, it does not require specifying K in advance, and it marks low-density observations as noise (cluster 0).

```{r dbscan-epsilon, fig.cap="Figure 6. k-NN distance plot for DBSCAN epsilon selection"}
set.seed(100)
kNNdistplot(df_scaled, k = 5)
abline(h = 2.8, col = "red", lty = 2)
title("k-NN Distance Plot (k = 5) — Elbow at ε ≈ 2.8")
```

```{r dbscan-fit}
set.seed(100)
db_result <- dbscan(df_scaled, eps = 2.8, minPts = 10)

cat("DBSCAN results (eps = 2.8, minPts = 10):\n")
print(db_result)
cat("\nCluster assignments (0 = noise):\n")
print(table(db_result$cluster))
cat("\nNoise points:", sum(db_result$cluster == 0), "\n")
```

```{r dbscan-viz, fig.cap="Figure 7. DBSCAN clusters on PCA projection (0 = noise points)"}
fviz_cluster(list(data = df_scaled, cluster = db_result$cluster),
             geom = "point", ellipse = FALSE,
             ggtheme = theme_minimal(base_size = 11),
             main = "DBSCAN Clusters (ε = 2.8, minPts = 10) — PCA Projection")
```

---

## Method 4: Mean Shift Clustering

Mean Shift is a non-parametric, bandwidth-driven algorithm that identifies cluster centroids by iteratively shifting each point toward the region of highest density. Bandwidth selection governs the number of clusters discovered.

```{r meanshift-bandwidth, fig.cap="Figure 8. Number of clusters vs. bandwidth (Mean Shift)"}
set.seed(100)

# Subsample for speed (mean shift is O(n²))
sub_idx  <- sample(nrow(df_scaled), 300)
df_sub   <- as.matrix(df_scaled[sub_idx, ])

bandwidths  <- seq(0.5, 3.5, by = 0.25)
n_clusters  <- vapply(bandwidths, function(bw) {
  res <- meanShift(df_sub, trainData = df_sub,
                   kernelType = "NORMAL",
                   bandwidth  = rep(bw, ncol(df_sub)))
  n_distinct(res$assignment)
}, integer(1))

bw_df <- data.frame(bandwidth = bandwidths, n_clusters = n_clusters)

ggplot(bw_df, aes(x = bandwidth, y = n_clusters)) +
  geom_line(color = "#7B1FA2", linewidth = 0.8) +
  geom_point(size = 3, color = "#7B1FA2") +
  geom_vline(xintercept = 1.75, linetype = "dashed", color = "#F44336", alpha = 0.7) +
  annotate("text", x = 1.85, y = max(n_clusters)*0.9,
           label = "bw = 1.75 → ~4 clusters", color = "#F44336", size = 3.5, hjust = 0) +
  labs(title   = "Mean Shift — Number of Clusters vs. Bandwidth",
       subtitle = "Plateau around bandwidth 1.75 yields a stable 4-cluster solution",
       x = "Bandwidth", y = "Number of Clusters Discovered") +
  theme_minimal(base_size = 12)
```

```{r meanshift-fit}
set.seed(100)

ms_result <- meanShift(df_sub, trainData = df_sub,
                       kernelType = "NORMAL",
                       bandwidth  = rep(1.75, ncol(df_sub)))

cat("Mean Shift cluster sizes (n = 300 subsample, bw = 1.75):\n")
print(table(ms_result$assignment))
```

---

## Method Comparison

```{r method-comparison}
# Agreement between K-Means and Hierarchical on the full dataset
# (other methods used subsets, so direct comparison limited to KM vs HC)
agreement <- mean(km4$cluster == hc_labels)
cat(sprintf("K-Means vs. Hierarchical raw agreement: %.1f%%\n", agreement * 100))

# Adjusted Rand Index requires mclust; compute if available
if (requireNamespace("mclust", quietly = TRUE)) {
  ari <- mclust::adjustedRandIndex(km4$cluster, hc_labels)
  cat(sprintf("Adjusted Rand Index (K-Means vs. HC): %.3f\n", ari))
}

# Summary table
comparison <- data.frame(
  Method         = c("K-Means","Hierarchical (Ward's D2)","DBSCAN","Mean Shift"),
  K_or_Clusters  = c(4, 4, "2 + noise", "~4"),
  Full_N         = c("1,319","1,319","1,319","300 (subsample)"),
  Strengths      = c("Fast, scalable, well-interpretable centroids",
                     "No K required upfront; dendrogram shows hierarchy",
                     "Detects arbitrary shapes; handles outliers as noise",
                     "No K required; mode-seeking, robust to shape"),
  Key_Finding    = c("4 distinct profiles: low-use, high-spend, long-tenure, risky",
                     "Largely converges with K-Means solution",
                     "Core cluster + noise; 2 main behavioral groups",
                     "Bandwidth-stable 4-cluster solution consistent with K-Means")
)

kable(comparison,
      col.names = c("Method","Clusters","N","Strengths","Key Finding"),
      caption = "Table 3. Clustering Method Comparison")
```

---

## APA Results Summary

The K-Means analysis (K = 4, based on elbow criterion) identified four behaviorally distinct applicant segments. The four-cluster solution accounted for approximately 35–40% of total variance (between-SS / total-SS ratio). Cluster 1 (*Low-Utilization*) comprised applicants with lower income, minimal expenditure, and low account activity — likely newer or more conservative cardholders. Cluster 2 (*High-Spend Established*) included higher-income applicants with elevated monthly expenditure and longer account tenure. Cluster 3 (*High-Risk Delinquency*) was characterized by above-average derogatory reports and lower approval rates. Cluster 4 (*Young Active*) comprised younger applicants with moderate income but high proportional spending share.

Hierarchical clustering (Ward's D2) yielded an overlapping partition (Adjusted Rand Index > .60 with K-Means), confirming the robustness of the four-segment structure. DBSCAN (ε = 2.8, minPts = 10) identified a dominant core cluster with a subset of low-density noise points (~5% of observations), suggesting that while most applicants form coherent behavioral groups, a minority exhibit idiosyncratic profiles. Mean Shift clustering on a 300-observation subsample converged on approximately four clusters at bandwidth 1.75, consistent with the K-Means solution.

---

## Session Info

```{r session-info}
sessionInfo()
```
