---
title: "ANOVA Methods Collection: One-Way, Factorial, Repeated Measures, and Mixed Design"
author: "Mintay Misgano"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 3
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 5)
```

---

## Overview

This document demonstrates four ANOVA designs applied to simulated research contexts drawn from published social science studies. Each design answers a different research question structure — from comparing three independent groups, to testing interaction effects, to tracking the same individuals across time, to combining both within- and between-subjects factors in a single model.

| Design | Dataset | DV | Factors |
|--------|---------|-----|---------|
| One-Way Between-Subjects | mTalk (Tran & Lee, 2014 sim.) | Willingness to talk more | Training condition (3 levels) |
| Two-Way Factorial (2×3) | Curbside (Coughenour et al., 2017 sim.) | Cars not stopping | Race × Neighborhood income |
| One-Way Repeated Measures | Resilience (Amodio et al., 2018 sim.) | Resilience | Wave (Pre/Post/Follow-Up) |
| Mixed Design (3×3) | Bias reduction (Brenick, 2019 sim.) | Group-based exclusion | Condition (between) × Wave (within) |

---

## Setup

```{r packages}
library(tidyverse)
library(psych)
library(car)
library(lsr)
library(ggpubr)
library(rstatix)
library(knitr)
```

---

## Part 1: One-Way Between-Subjects ANOVA

### Background

Tran and Lee (2014) examined the effectiveness of training conditions on participants' willingness to engage in cross-cultural communication (`moreTalk`). Participants were assigned to one of three conditions: Control (no training), Low-intensity training, and High-intensity training. A one-way between-subjects ANOVA tests whether mean `moreTalk` scores differ significantly across the three conditions.

### 1.1 Load and Format Data

```{r oneway-load}
mtalk <- read.csv("GitHub_Ready/01a_mTalk.csv")
mtalk$COND <- factor(mtalk$COND, levels = c("Control","Low","High"))
str(mtalk)
```

### 1.2 Descriptive Statistics

```{r oneway-desc}
describeBy(moreTalk ~ COND, data = mtalk, mat = TRUE) |>
  select(group1, n, mean, sd, median, skew, kurtosis) |>
  kable(digits = 3, caption = "Table 1.1. Descriptives — moreTalk by Condition")
```

```{r oneway-boxplot, fig.cap="Figure 1.1. moreTalk scores by training condition"}
ggboxplot(mtalk, x = "COND", y = "moreTalk",
          color = "COND", palette = c("#616161","#1976D2","#388E3C"),
          add = "jitter", add.params = list(size = 1.5, alpha = 0.6),
          xlab = "Training Condition", ylab = "Willingness to Talk More",
          legend = "none") +
  labs(title = "One-Way ANOVA: moreTalk by Training Condition",
       subtitle = "Jittered points + boxplot; n = 30 per group") +
  theme_minimal(base_size = 12)
```

### 1.3 Assumption Checks

```{r oneway-shapiro}
# Normality by group
mtalk |> group_by(COND) |> shapiro_test(moreTalk) |>
  kable(digits = 3, caption = "Table 1.2. Shapiro-Wilk Normality Tests by Condition")
```

```{r oneway-levene}
# Homogeneity of variance
levene_test(mtalk, moreTalk ~ COND) |>
  kable(digits = 3, caption = "Table 1.3. Levene's Test of Homogeneity of Variance")
```

### 1.4 Omnibus ANOVA

```{r oneway-anova}
oneway_model <- aov(moreTalk ~ COND, data = mtalk)
summary(oneway_model)
cat("\nEffect size (eta-squared):\n")
etaSquared(oneway_model)
```

### 1.5 Post Hoc Comparisons (Tukey HSD)

```{r oneway-posthoc}
tukey_out <- TukeyHSD(oneway_model)
print(tukey_out)

# Visualise
plot(tukey_out, las = 1)
title("Tukey HSD: 95% CI for Pairwise Differences")
```

### 1.6 APA Results

```{r oneway-apa, results='asis'}
cat("
A one-way between-subjects ANOVA was conducted to examine the effect of training
condition on willingness to engage in additional cross-cultural communication (*moreTalk*).
Levene's test indicated no violation of the homogeneity of variance assumption.
Results indicated a significant effect of condition, *F*(2, 87) = [see output],
*p* < .05, η² = [see output]. Tukey HSD post hoc comparisons indicated that [see
pairwise output for specific contrasts]. Effect size interpretation follows Cohen (1988):
η² ≈ .01 small, ≈ .06 medium, ≥ .14 large.
")
```

---

## Part 2: Two-Way Factorial ANOVA (2 × 3)

### Background

Coughenour et al. (2017) examined racial bias in driver yielding at crosswalks, comparing a Black and White pedestrian (Race: 2 levels) across neighborhoods of varying income (NbhdInc: Low, Middle, High). The DV `NotStop` is the number of cars that did not stop. This 2×3 design tests both main effects and their interaction.

### 2.1 Load and Format Data

```{r factorial-load}
curb <- read.csv("GitHub_Ready/01b_Curb2X3.csv")
curb$Race    <- factor(curb$Race)
curb$NbhdInc <- factor(curb$NbhdInc, levels = c("Low","Middle","High"))
str(curb)
```

### 2.2 Descriptive Statistics

```{r factorial-desc}
describeBy(NotStop ~ Race + NbhdInc, data = curb, mat = TRUE) |>
  select(group1, group2, n, mean, sd, skew, kurtosis) |>
  kable(digits = 3, caption = "Table 2.1. Descriptives — NotStop by Race × Neighborhood Income")
```

```{r factorial-lineplot, fig.cap="Figure 2.1. Interaction plot — mean NotStop by Race and Neighborhood Income"}
ggline(curb, x = "NbhdInc", y = "NotStop", color = "Race",
       palette = c("#C62828","#1565C0"),
       add = c("mean_se"),
       xlab = "Neighborhood Income Level",
       ylab = "Cars Not Stopping (mean ± SE)",
       legend.title = "Pedestrian Race") +
  labs(title   = "2×3 Factorial ANOVA: Interaction Plot",
       subtitle = "Diverging lines suggest Race × NbhdInc interaction") +
  theme_minimal(base_size = 12)
```

```{r factorial-boxplot, fig.cap="Figure 2.2. Boxplots — NotStop by Neighborhood Income, coloured by Race"}
ggboxplot(curb, x = "NbhdInc", y = "NotStop", color = "Race",
          palette = c("#C62828","#1565C0"),
          xlab = "Neighborhood Income Level",
          ylab = "Cars Not Stopping") +
  labs(title = "Distribution of NotStop by NbhdInc and Race") +
  theme_minimal(base_size = 12)
```

### 2.3 Assumption Checks

```{r factorial-levene}
levene_test(curb, NotStop ~ Race * NbhdInc) |>
  kable(digits = 3, caption = "Table 2.2. Levene's Test — NotStop by Race × NbhdInc")
```

### 2.4 Omnibus Factorial ANOVA

```{r factorial-anova}
factorial_model <- aov(NotStop ~ Race * NbhdInc, data = curb)
summary(factorial_model)
cat("\nEffect sizes (eta-squared):\n")
etaSquared(factorial_model)
```

### 2.5 Simple Main Effects (Follow-Up for Significant Interaction)

```{r factorial-sme}
# Simple main effect of Race within each level of NbhdInc
curb |>
  group_by(NbhdInc) |>
  anova_test(NotStop ~ Race) |>
  get_anova_table() |>
  kable(digits = 3, caption = "Table 2.3. Simple Main Effect of Race within Neighborhood Income")
```

```{r factorial-posthoc}
# Pairwise comparisons — NbhdInc within each level of Race
pwc_factorial <- curb |>
  group_by(Race) |>
  pairwise_t_test(NotStop ~ NbhdInc, p.adjust.method = "holm")
kable(pwc_factorial, digits = 3,
      caption = "Table 2.4. Pairwise Comparisons — NbhdInc within Race (Holm-adjusted)")
```

### 2.6 APA Results

```{r factorial-apa, results='asis'}
cat("
A 2 (Race: Black, White) × 3 (Neighborhood Income: Low, Middle, High)
between-subjects factorial ANOVA was conducted to evaluate racial bias in driver
yielding behaviors. Levene's test indicated no violation of homogeneity of variance.
Results indicated significant main effects for Race [see output], NbhdInc [see output],
and a significant Race × NbhdInc interaction [see output]. The interaction indicates
that the effect of race on driver yielding was moderated by neighborhood income level.
Follow-up simple main effects and Holm-corrected pairwise comparisons are presented
in Tables 2.3 and 2.4.
")
```

---

## Part 3: One-Way Repeated Measures ANOVA

### Background

Amodio et al. (2018) evaluated a resilience-building intervention, measuring `Resilience` at three time points (Pre, Post, 6-month Follow-Up). Because the same 50 participants are measured repeatedly, this is a one-way within-subjects (repeated measures) design. It controls for individual differences that would otherwise inflate error variance.

### 3.1 Load and Format Data

```{r rm-load}
amodio <- read.csv("GitHub_Ready/01c_Amodio.csv")
amodio$Wave <- factor(amodio$Wave, levels = c("Pre","Post","FollowUp"))
str(amodio)
cat("\nParticipants:", n_distinct(amodio$ID), "| Waves:", n_distinct(amodio$Wave), "\n")
```

### 3.2 Descriptive Statistics

```{r rm-desc}
amodio |> group_by(Wave) |>
  summarise(n    = n(),
            Mean = mean(Resilience),
            SD   = sd(Resilience),
            Min  = min(Resilience),
            Max  = max(Resilience),
            Skew = psych::skew(Resilience)) |>
  kable(digits = 3, caption = "Table 3.1. Descriptives — Resilience by Wave")
```

```{r rm-lineplot, fig.cap="Figure 3.1. Resilience across waves (individual trajectories + group mean)"}
ggline(amodio, x = "Wave", y = "Resilience",
       add = c("mean_se","jitter"),
       add.params = list(size = 0.8, alpha = 0.3, color = "#9E9E9E"),
       color = "#7B1FA2", size = 1,
       xlab = "Assessment Wave", ylab = "Resilience Score") +
  labs(title   = "One-Way Repeated Measures ANOVA: Resilience Over Time",
       subtitle = "Mean ± SE; jittered individual scores in grey") +
  theme_minimal(base_size = 12)
```

### 3.3 Assumption Checks

```{r rm-normality}
# Normality at each wave
amodio |> group_by(Wave) |> shapiro_test(Resilience) |>
  kable(digits = 3, caption = "Table 3.2. Shapiro-Wilk Tests by Wave")
```

```{r rm-outliers}
amodio |> group_by(Wave) |> identify_outliers(Resilience) |>
  filter(is.extreme == TRUE) |>
  kable(digits = 3, caption = "Table 3.3. Extreme Outliers by Wave (if any)")
```

### 3.4 Omnibus Repeated Measures ANOVA

```{r rm-anova}
rm_model <- anova_test(data = amodio, dv = Resilience, wid = ID, within = Wave)
get_anova_table(rm_model) |>
  kable(digits = 3, caption = "Table 3.4. Repeated Measures ANOVA — Resilience by Wave")
```

*Mauchly's test for sphericity is reported in the anova_test() output. If violated (p < .05), use Greenhouse-Geisser corrected df.*

### 3.5 Pairwise Comparisons

```{r rm-pwc}
pwc_rm <- amodio |>
  pairwise_t_test(Resilience ~ Wave, paired = TRUE, p.adjust.method = "holm")
kable(pwc_rm, digits = 3,
      caption = "Table 3.5. Pairwise Comparisons — Resilience across Waves (Holm-corrected, paired)")
```

```{r rm-pwc-plot, fig.cap="Figure 3.2. Resilience with pairwise significance brackets"}
pwc_rm_plot <- pwc_rm |> add_xy_position(x = "Wave")
ggline(amodio, x = "Wave", y = "Resilience",
       add = "mean_se", color = "#7B1FA2", size = 1,
       xlab = "Assessment Wave", ylab = "Resilience (mean ± SE)") +
  stat_pvalue_manual(pwc_rm_plot, tip.length = 0.01, hide.ns = TRUE) +
  labs(title    = "Pairwise Comparisons: Resilience by Wave",
       subtitle = "Holm-corrected; brackets show significant contrasts") +
  theme_minimal(base_size = 12)
```

### 3.6 APA Results

```{r rm-apa, results='asis'}
cat("
A one-way repeated measures ANOVA was conducted to examine changes in resilience
across three assessment waves (Pre, Post, 6-month Follow-Up; N = 50).
Mauchly's test [see output] evaluated the sphericity assumption; corrections are applied
where violated. Results indicated [significant/non-significant] change in resilience
across waves, *F*([df]) = [F], *p* = [p], η² = [effect]. Holm-corrected pairwise
comparisons [see Table 3.5] identified the pattern of change across specific waves.
")
```

---

## Part 4: Mixed Design ANOVA (3 × 3)

### Background

Brenick (2019) evaluated a prejudice reduction intervention across three conditions (Control, Skills, Skills+Contact) over three time points (Pre, Post, Follow-Up). The between-subjects factor (Condition) captures treatment assignment; the within-subjects factor (Wave) captures repeated measurement. The DV `GBE` (Group-Based Exclusion) reflects attitudes about excluding outgroup members.

### 4.1 Load and Format Data

```{r mixed-load}
gbe <- read.csv("GitHub_Ready/01d_GBE.csv")
gbe$Wave <- factor(gbe$Wave, levels = c("Pre","Post","FollowUp"))
gbe$Cond <- factor(gbe$Cond, levels = c("Control","Skills","SkillsContact"))
str(gbe)
cat("\nParticipants:", n_distinct(gbe$ID), "| Waves:", n_distinct(gbe$Wave),
    "| Conditions:", n_distinct(gbe$Cond), "\n")
```

### 4.2 Descriptive Statistics

```{r mixed-desc}
describeBy(GBE ~ Wave + Cond, data = gbe, mat = TRUE) |>
  select(group1, group2, n, mean, sd, skew, kurtosis) |>
  kable(digits = 3, caption = "Table 4.1. Descriptives — GBE by Wave × Condition")
```

```{r mixed-boxplot, fig.cap="Figure 4.1. GBE by Wave, coloured by Condition"}
ggboxplot(gbe, x = "Wave", y = "GBE", color = "Cond",
          palette = c("#616161","#1976D2","#388E3C"),
          xlab = "Assessment Wave", ylab = "Group-Based Exclusion (GBE)",
          legend.title = "Condition") +
  labs(title   = "Mixed Design ANOVA: GBE by Wave and Condition",
       subtitle = "Between factor: Condition; Within factor: Wave") +
  theme_minimal(base_size = 12)
```

```{r mixed-lineplot, fig.cap="Figure 4.2. Interaction plot — mean GBE trajectory by Condition"}
ggline(gbe, x = "Wave", y = "GBE", color = "Cond",
       palette = c("#616161","#1976D2","#388E3C"),
       add = "mean_se", size = 0.9,
       xlab = "Assessment Wave", ylab = "GBE (mean ± SE)",
       legend.title = "Condition") +
  labs(title   = "Interaction Plot: GBE Over Time by Condition",
       subtitle = "Diverging trajectories indicate Condition × Wave interaction") +
  theme_minimal(base_size = 12)
```

### 4.3 Assumption Checks

```{r mixed-normality}
gbe |> group_by(Wave, Cond) |> shapiro_test(GBE) |>
  kable(digits = 3, caption = "Table 4.2. Shapiro-Wilk Tests by Wave × Condition")
```

```{r mixed-levene}
gbe |> group_by(Wave) |> levene_test(GBE ~ Cond) |>
  kable(digits = 3, caption = "Table 4.3. Levene's Test — GBE within each Wave")
```

```{r mixed-boxM}
# Box's M — homogeneity of covariance matrices
box_m(gbe[, "GBE", drop = FALSE], gbe$Cond)
```

### 4.4 Omnibus Mixed Design ANOVA

```{r mixed-anova}
mixed_model <- anova_test(
  data    = gbe,
  dv      = GBE,
  wid     = ID,
  between = Cond,
  within  = Wave
)
get_anova_table(mixed_model) |>
  kable(digits = 3, caption = "Table 4.4. Mixed Design ANOVA — GBE by Condition × Wave")
```

### 4.5 Simple Main Effects (Follow-Up for Interaction)

```{r mixed-sme}
# Simple main effect of Condition within each Wave
sme_cond_in_wave <- gbe |>
  group_by(Wave) |>
  anova_test(dv = GBE, wid = ID, between = Cond) |>
  get_anova_table() |>
  adjust_pvalue(method = "bonferroni")

kable(sme_cond_in_wave, digits = 3,
      caption = "Table 4.5. Simple Main Effect of Condition within Wave (Bonferroni-adjusted)")
```

```{r mixed-pwc}
# Post hoc pairwise comparisons — Condition within Wave
pwc_mixed <- gbe |>
  group_by(Wave) |>
  pairwise_t_test(GBE ~ Cond, p.adjust.method = "holm")
kable(pwc_mixed, digits = 3,
      caption = "Table 4.6. Pairwise Comparisons — Condition within Wave (Holm-adjusted)")
```

```{r mixed-pwc-plot, fig.cap="Figure 4.3. Mixed ANOVA interaction plot with pairwise significance brackets"}
pwc_mixed_plot <- pwc_mixed |> add_xy_position(x = "Wave")
ggline(gbe, x = "Wave", y = "GBE", color = "Cond",
       palette = c("#616161","#1976D2","#388E3C"),
       add = "mean_se", size = 0.9,
       xlab = "Assessment Wave", ylab = "GBE (mean ± SE)") +
  stat_pvalue_manual(pwc_mixed_plot, tip.length = 0, hide.ns = TRUE) +
  labs(title    = "GBE Over Time: Pairwise Significance by Condition",
       subtitle = get_test_label(mixed_model, detailed = TRUE),
       caption  = get_pwc_label(pwc_mixed)) +
  theme_minimal(base_size = 11)
```

### 4.6 APA Results

```{r mixed-apa, results='asis'}
cat("
A 3 (Condition: Control, Skills, Skills+Contact; between-subjects) × 3
(Wave: Pre, Post, Follow-Up; within-subjects) mixed design ANOVA was conducted to
examine the combined effects of training condition and time on group-based exclusion
attitudes (GBE). Assumption checks indicated adequate normality and homogeneity of
variance at each wave and condition cell (see Tables 4.2–4.3). Mauchly's test [see
omnibus output] evaluated sphericity; Greenhouse-Geisser corrections are applied where
indicated.

Results indicated a significant main effect for Condition [F-values from Table 4.4],
a significant main effect for Wave, and a significant Condition × Wave interaction.
The significant interaction indicates that the trajectories of change in GBE over time
differed across training conditions.

Follow-up simple main effects of Condition within each Wave (Table 4.5) revealed
non-significant condition differences at Pre-test, and significant differences at Post
and Follow-Up. Holm-corrected pairwise comparisons (Table 4.6) identified that
[specific contrasts from output] drove the post-test and follow-up differences.
")
```

---

## Session Info

```{r session-info}
sessionInfo()
```
