---
title: "Measurement Invariance Testing: Ableist Microaggressions Scale"
author: "Mintay Misgano"
date: "2026-04-06"
output:
  github_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.path = "03_Analysis_Workflow_files/figure-gfm/"
)
```

## Overview

This workflow tests whether the Ableist Microaggressions Scale (AMS) keeps the same measurement structure across mild and severe disability-severity groups. The sequence follows the standard measurement-invariance progression: configural, weak, strong, and strict. The analysis uses lavaan for the constrained multi-group models because lavaan supports the `group.equal` constraints needed for formal invariance testing.

## Setup

```{r libraries}
# Load the psychometric modeling and reporting packages used in the workflow.
# lavaan estimates the multi-group CFA models, semPlot renders the model
# structure, and the table packages keep the rendered workflow readable.
library(lavaan)
library(semPlot)
library(psych)
library(MASS)
library(tidyverse)
library(knitr)
```

## Setup and Data Simulation

```{r simulate-data}
# Recreate the public simulated AMS dataset from the published loading matrix.
# This mirrors the rendered workflow and keeps the data-generation assumptions
# visible: standardized factors, item-specific uniqueness, integer 0-5 response
# scaling, and severity grouping from ranked AMS mean scores.
set.seed(211023)

AMS_Imat <- matrix(c(
  .74, .75, .65, .58, .62,
  .01, .05, -.08,
  .00, .03, .01, .04, .25,
  -.06, -.02, .11, .18, .25, .26, .14,
  -.03, .00, .20, -.07, .15,
  .71, .52, .47,
  .02, .04, .00, -.01, .01,
  -.18, .07, .14, -.17, .05, -.12, .16,
  .11, -.07, -.03, .20, .03,
  .00, .07, .15,
  .91, .85, .64, .56, .42,
  .04, .04, -.15, .03, .13, .07, .14,
  -.12, .06, .16, -.01, .02,
  -.07, .05, .20,
  -.01, .01, .19, .16, .21,
  .89, .73, .70, .46, .41, .40, .32
), ncol = 4)

item_names <- c(
  paste0("Help", 1:5),
  paste0("Min", 1:3),
  paste0("Pers", 1:5),
  paste0("Oth", 1:7)
)

factor_names <- c(
  "Helplessness",
  "Minimization",
  "DenialPersonhood",
  "Otherization"
)

rownames(AMS_Imat) <- item_names
colnames(AMS_Imat) <- factor_names

communalities <- rowSums(AMS_Imat^2)
uniqueness <- 1 - communalities
R_implied <- AMS_Imat %*% t(AMS_Imat) + diag(uniqueness)

raw_data <- MASS::mvrnorm(
  n = 833,
  mu = rep(2, 20),
  Sigma = R_implied,
  empirical = TRUE
)

colnames(raw_data) <- item_names

dfAMSi <- as.data.frame(
  apply(
    raw_data,
    2,
    function(x) as.integer(pmin(pmax(round(x), 0), 5))
  )
)

dfAMSi$AMSm_total <- rowMeans(dfAMSi)
dfAMSi$Severity <- ifelse(
  rank(dfAMSi$AMSm_total, ties.method = "first") <= 548,
  "Mild",
  "Severe"
)
dfAMSi$AMSm_total <- NULL

table(dfAMSi$Severity)
```

## Descriptive Checks

```{r descriptive-statistics}
# Summarize item-level endorsement by severity group before modeling. This
# descriptive step is not the final decision; it identifies the group-mean
# pattern that the invariance models need to evaluate.
mild_desc <- psych::describe(dfAMSi[dfAMSi$Severity == "Mild", item_names])
severe_desc <- psych::describe(dfAMSi[dfAMSi$Severity == "Severe", item_names])

round(mild_desc, 2) |>
  knitr::kable(caption = "Descriptive Statistics - Mild Group")

round(severe_desc, 2) |>
  knitr::kable(caption = "Descriptive Statistics - Severe Group")
```

## CFA Model

```{r model-specification}
# The four-factor model mirrors the published AMS structure. Keeping the model
# specification explicit makes it clear which items define each latent domain.
AMS4CorrMod <- '
  Helplessness     =~ Help1 + Help2 + Help3 + Help4 + Help5
  Minimization     =~ Min1  + Min2  + Min3
  DenialPersonhood =~ Pers1 + Pers2 + Pers3 + Pers4 + Pers5
  Otherization     =~ Oth1  + Oth2  + Oth3  + Oth4  + Oth5  + Oth6 + Oth7
'
```

## Invariance Models

```{r fit-models}
# Fit increasingly constrained multi-group CFA models. Each step adds one
# equality constraint family so the fit comparison can isolate which type of
# equivalence is supported by the data.
fit_configural <- lavaan::cfa(
  AMS4CorrMod,
  data = dfAMSi,
  group = "Severity",
  estimator = "ML"
)

fit_weak <- lavaan::cfa(
  AMS4CorrMod,
  data = dfAMSi,
  group = "Severity",
  estimator = "ML",
  group.equal = c("loadings")
)

fit_strong <- lavaan::cfa(
  AMS4CorrMod,
  data = dfAMSi,
  group = "Severity",
  estimator = "ML",
  group.equal = c("loadings", "intercepts")
)

fit_strict <- lavaan::cfa(
  AMS4CorrMod,
  data = dfAMSi,
  group = "Severity",
  estimator = "ML",
  group.equal = c("loadings", "intercepts", "residuals")
)
```

## Fit Tables

```{r fit-index-summary}
# Extract a compact fit table that can be checked against the report. The
# report interprets these values; this workflow preserves the reproducible
# calculation chain.
extract_fit <- function(fit, model_name) {
  fi <- lavaan::fitMeasures(
    fit,
    c(
      "chisq", "df", "pvalue", "cfi", "tli",
      "rmsea", "rmsea.ci.lower", "rmsea.ci.upper",
      "srmr", "aic", "bic"
    )
  )

  data.frame(
    Model = model_name,
    chi2 = round(fi["chisq"], 3),
    df = fi["df"],
    p = round(fi["pvalue"], 3),
    CFI = round(fi["cfi"], 3),
    TLI = round(fi["tli"], 3),
    RMSEA = round(fi["rmsea"], 3),
    RMSEA_CI = paste0(
      "[",
      round(fi["rmsea.ci.lower"], 3),
      ", ",
      round(fi["rmsea.ci.upper"], 3),
      "]"
    ),
    SRMR = round(fi["srmr"], 3),
    AIC = round(fi["aic"], 2),
    BIC = round(fi["bic"], 2)
  )
}

fit_table <- rbind(
  extract_fit(fit_configural, "Configural"),
  extract_fit(fit_weak, "Weak"),
  extract_fit(fit_strong, "Strong"),
  extract_fit(fit_strict, "Strict diagnostic")
)

knitr::kable(
  fit_table,
  col.names = c(
    "Model", "chi-square", "df", "p", "CFI", "TLI", "RMSEA",
    "90% CI", "SRMR", "AIC", "BIC"
  ),
  caption = "Fit Indices for Measurement Invariance Sequence - AMS"
)
```

```{r sequential-comparison}
# Compare adjacent models. Weak invariance is supported when loading equality
# does not meaningfully degrade fit; strong invariance is rejected here because
# intercept equality produces a large practical fit decline.
lrt_all <- lavaan::lavTestLRT(fit_configural, fit_weak, fit_strong, fit_strict)
lrt_all

cfi_vals <- sapply(
  list(fit_configural, fit_weak, fit_strong, fit_strict),
  function(fit) lavaan::fitMeasures(fit, "cfi")
)

chi_vals <- sapply(
  list(fit_configural, fit_weak, fit_strong, fit_strict),
  function(fit) lavaan::fitMeasures(fit, "chisq")
)

df_vals <- sapply(
  list(fit_configural, fit_weak, fit_strong, fit_strict),
  function(fit) lavaan::fitMeasures(fit, "df")
)

delta_table <- data.frame(
  Comparison = c(
    "-",
    "Weak vs. configural",
    "Strong vs. weak",
    "Strict vs. strong"
  ),
  Delta_chi2 = c(NA, diff(chi_vals)),
  Delta_df = c(NA, diff(df_vals)),
  Delta_CFI = c(NA, diff(cfi_vals)),
  Decision = c("-", "Supported", "Not supported", "Diagnostic only")
)

knitr::kable(
  delta_table,
  col.names = c("Comparison", "Delta chi-square", "Delta df", "Delta CFI", "Decision"),
  digits = 3,
  caption = "Sequential Model Comparison"
)
```

## Configural Model Figure

```{r configural-path, fig.cap="Configural model path diagrams by severity group"}
# Render the configural structure as a visual check. The path figures are not
# the final evidence for invariance, but they help readers see the four-factor
# model being tested.
semPlot::semPaths(
  fit_configural,
  what = "std",
  layout = "tree",
  style = "lisrel",
  residuals = FALSE,
  intercepts = FALSE,
  nCharNodes = 0,
  edge.label.cex = 0.55,
  sizeMan = 4,
  sizeLat = 6,
  title = FALSE
)
```

## APA Results Summary

```{r apa-summary}
# Print a concise interpretation using the same values reported in the public
# report. The Delta chi-square degrees of freedom use the difference df from
# lavTestLRT, not the full model df.
lrt_weak <- lavaan::lavTestLRT(fit_configural, fit_weak)
lrt_strong <- lavaan::lavTestLRT(fit_weak, fit_strong)

cat(sprintf(
  paste(
    "Configural fit: chi-square(%d) = %.3f, CFI = %.3f, RMSEA = %.3f, SRMR = %.3f.\\n",
    "Weak vs. configural: Delta chi-square(%d) = %.3f, p = %.3f, Delta CFI = %.3f.\\n",
    "Strong vs. weak: Delta chi-square(%d) = %.3f, p < .001, Delta CFI = %.3f.\\n",
    "Conclusion: weak invariance is supported, but strong invariance is not supported."
  ),
  lavaan::fitMeasures(fit_configural, "df"),
  lavaan::fitMeasures(fit_configural, "chisq"),
  lavaan::fitMeasures(fit_configural, "cfi"),
  lavaan::fitMeasures(fit_configural, "rmsea"),
  lavaan::fitMeasures(fit_configural, "srmr"),
  as.integer(lrt_weak$`Df diff`[2]),
  lrt_weak$`Chisq diff`[2],
  lrt_weak$`Pr(>Chisq)`[2],
  abs(cfi_vals[2] - cfi_vals[1]),
  as.integer(lrt_strong$`Df diff`[2]),
  lrt_strong$`Chisq diff`[2],
  abs(cfi_vals[3] - cfi_vals[2])
))
```

## Session Info

```{r session-info}
sessionInfo()
```
