---
title: "Survival, Group, and Regression Analysis of Advanced Lung Cancer Outcomes"
subtitle: "An R statistical-methods workflow on the NCCTG lung cancer dataset (N = 228)"
author: "Mintay Misgano, PhD"
date: "2023"
output:
  github_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 5, dpi = 150
)
```

> **Question framing.** Among patients with advanced lung cancer, *who survives longer, and why?* This workflow answers that with one dataset and a sequence of methods chosen to match the structure of the outcome: group comparisons for "is there a difference," regression for "how do predictors relate to outcome," and survival analysis for "how long until the event, accounting for patients still alive at last contact (censoring)."

# 1. Setup and Data

The data are the NCCTG lung cancer records distributed with R's `survival` package (Loprinzi et al., 1994): 228 patients with advanced lung cancer, each with a survival time in days, a censoring indicator, and clinical predictors (age, sex, ECOG and Karnofsky performance scores, weight loss, calorie intake).

```{r load-packages}
library(survival)   # lung data, survfit (Kaplan-Meier), coxph (Cox model), cox.zph (PH check)
library(dplyr)      # data manipulation
library(tidyr)      # reshaping for plots
library(broom)      # tidy model output into data frames for ggplot
library(ggplot2)    # visualization
library(scales)     # axis formatting
```

```{r prep-data}
# Load the lung dataset and build analysis-ready variables.
# In survival::lung, status is coded 1 = censored, 2 = dead. We recode an explicit
# event indicator (1 = death observed, 0 = censored) so every model reads the outcome
# the same way, and we label sex so plots and tables are self-explanatory.
data(lung)
d <- lung %>%
  mutate(
    event   = ifelse(status == 2, 1, 0),                 # 1 = death, 0 = censored
    sex_lab = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
    ecog    = factor(ph.ecog)                            # ECOG performance status as a group factor
  )

# Cohort overview: how many patients, how many deaths vs. censored, basic age profile.
c(n = nrow(d), deaths = sum(d$event), censored = sum(d$event == 0))
summary(d$age)
table(d$sex_lab)
```

# 2. Exploratory Description

```{r eda-survival-time}
# Distribution of observed follow-up time, split by whether the patient died or was
# censored. Censored observations (still alive at last contact) carry partial information
# and are exactly why ordinary averages of "time" would be misleading here.
ggplot(d, aes(x = time, fill = factor(event, labels = c("Censored", "Died")))) +
  geom_histogram(bins = 30, position = "stack", color = "white") +
  scale_fill_manual(values = c("Censored" = "#9ecae1", "Died" = "#08519c"), name = NULL) +
  labs(title = "Observed follow-up time by outcome",
       x = "Time (days)", y = "Patients") +
  theme_minimal()
```

```{r eda-status-by-sex}
# Event vs. censoring counts by sex. This previews the survival comparison in Section 3:
# a higher share of censored (surviving) patients in one group hints at better prognosis.
d %>%
  count(sex_lab, outcome = factor(event, labels = c("Censored", "Died"))) %>%
  ggplot(aes(x = sex_lab, y = n, fill = outcome)) +
  geom_col(position = "dodge", color = "white") +
  scale_fill_manual(values = c("Censored" = "#9ecae1", "Died" = "#08519c"), name = NULL) +
  labs(title = "Outcome counts by sex", x = NULL, y = "Patients") +
  theme_minimal()
```

# 3. Survival Analysis

## 3.1 Kaplan-Meier estimation

The Kaplan-Meier estimator (Kaplan & Meier, 1958) is a nonparametric estimate of the survival function S(t) = P(survival beyond time t). It steps down at each observed death and correctly keeps censored patients in the risk set until the moment they are censored, so it uses partial information instead of discarding it.

```{r km-overall}
# Overall survival curve with 95% confidence band. The median survival time is where
# the curve crosses 0.5.
fit_all <- survfit(Surv(time, event) ~ 1, data = d)
km_all  <- broom::tidy(fit_all)
ggplot(km_all, aes(x = time, y = estimate)) +
  geom_step(color = "#08519c") +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15, fill = "#08519c") +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
  labs(title = "Overall Kaplan-Meier survival", x = "Time (days)", y = "Survival probability") +
  theme_minimal()
```

```{r km-by-sex}
# Survival by sex. The gap between the two step functions is the effect we formally
# test with the log-rank statistic below.
fit_sex <- survfit(Surv(time, event) ~ sex_lab, data = d)
km_sex  <- broom::tidy(fit_sex)
ggplot(km_sex, aes(x = time, y = estimate, color = strata, fill = strata)) +
  geom_step() +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.12, color = NA) +
  scale_color_manual(values = c("sex_lab=Male" = "#08519c", "sex_lab=Female" = "#a50f15"),
                     labels = c("Female", "Male"), name = NULL) +
  scale_fill_manual(values = c("sex_lab=Male" = "#08519c", "sex_lab=Female" = "#a50f15"),
                    labels = c("Female", "Male"), name = NULL) +
  labs(title = "Survival by sex", x = "Time (days)", y = "Survival probability") +
  theme_minimal()

# Median survival per group and the log-rank test of the difference.
fit_sex
survdiff(Surv(time, event) ~ sex_lab, data = d)
```

# 4. Group Comparison: Performance Status and Survival Time

One-way ANOVA (Fisher, 1925) tests whether mean survival time differs across ECOG performance-status groups (0 = fully active to 3 = largely bedbound). ANOVA partitions total variance into between-group and within-group components; the F-statistic is their ratio.

```{r box-time-by-ecog}
# Survival time across ECOG performance levels. ANOVA tests the group-mean difference;
# the boxplot shows the direction and spread behind that test.
d_ec <- d %>% filter(!is.na(ph.ecog))
ggplot(d_ec, aes(x = ecog, y = time, fill = ecog)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.4) +
  scale_fill_brewer(palette = "Blues", guide = "none") +
  labs(title = "Survival time by ECOG performance status",
       x = "ECOG performance status", y = "Time (days)") +
  theme_minimal()

aov_fit <- aov(time ~ ecog, data = d_ec)
summary(aov_fit)
```

# 5. Multivariable Cox Proportional-Hazards Model

The Cox proportional-hazards model (Cox, 1972) relates predictors to the hazard (instantaneous risk of death) without assuming a shape for the baseline hazard. Each coefficient, exponentiated, is a hazard ratio (HR): HR > 1 means higher risk, HR < 1 means protection. We chose Cox over fully parametric survival models (e.g., Weibull) because it makes no distributional assumption about baseline risk, and over logistic regression because it uses *time-to-event* and censoring rather than collapsing the outcome to a yes/no.

```{r cox-forest}
# Multivariable Cox model. We tidy the fit into hazard ratios with 95% CIs and show them
# as a forest plot: points right of 1 raise risk, left of 1 lower it; whiskers crossing 1
# are not statistically distinguishable from "no effect."
cox_fit <- coxph(Surv(time, event) ~ age + sex_lab + ph.ecog + wt.loss, data = d)
hr <- broom::tidy(cox_fit, exponentiate = TRUE, conf.int = TRUE)
ggplot(hr, aes(x = estimate, y = reorder(term, estimate))) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  geom_point(size = 2.5, color = "#08519c") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2, color = "#08519c") +
  scale_x_log10() +
  labs(title = "Cox hazard ratios (95% CI)", x = "Hazard ratio (log scale)", y = NULL) +
  theme_minimal()

summary(cox_fit)            # coefficients, HRs, CIs, p-values, concordance (C-index)
```

```{r cox-ph-check}
# Proportional-hazards assumption check (Grambsch & Therneau, 1994): a non-significant
# global test means hazard ratios are reasonably constant over time, which the Cox model
# requires. Schoenfeld residuals should show no trend against time.
zph <- cox.zph(cox_fit)
zph
plot(zph[2], main = "Schoenfeld residuals: sex")
```

# 6. References

- Cox, D. R. (1972). Regression models and life-tables. *Journal of the Royal Statistical Society: Series B, 34*(2), 187-202.
- Fisher, R. A. (1925). *Statistical methods for research workers.* Oliver and Boyd.
- Grambsch, P. M., & Therneau, T. M. (1994). Proportional hazards tests and diagnostics based on weighted residuals. *Biometrika, 81*(3), 515-526.
- Harrell, F. E., Lee, K. L., & Mark, D. B. (1996). Multivariable prognostic models. *Statistics in Medicine, 15*(4), 361-387.
- Kaplan, E. L., & Meier, P. (1958). Nonparametric estimation from incomplete observations. *Journal of the American Statistical Association, 53*(282), 457-481.
- Loprinzi, C. L., et al. (1994). Prospective evaluation of prognostic variables from patient-completed questionnaires. *Journal of Clinical Oncology, 12*(3), 601-607.
