---
title: "Public-Safe Consulting Bid Accuracy Analysis"
subtitle: "Synthetic portfolio case study"
author: "Mintay Misgano, PhD"
output:
  github_document:
    toc: true
    toc_depth: 3
---

> This public portfolio version uses fully synthetic data. It does not contain real client records, personnel names, customer IDs, invoice numbers, contact fields, exact dates, or operational notes.

# 1. Setup

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

# 2. Load Public Data

```{r load-data}
df <- read.csv("data.csv", stringsAsFactors = FALSE)

df <- df %>%
  mutate(
    consultant_id = factor(consultant_id),
    client_id = factor(client_id),
    service_area = factor(service_area),
    project_type = factor(project_type),
    department_owner = factor(department_owner),
    associate_assigned = factor(associate_assigned),
    billable_travel = factor(billable_travel),
    billable_shipping = factor(billable_shipping)
  )

glimpse(df)
```

# 3. Outcome Definition

```{r outcome-check}
summary(df$bid_discrepancy)

df %>%
  summarise(
    records = n(),
    mean_discrepancy = mean(bid_discrepancy),
    median_discrepancy = median(bid_discrepancy),
    sd_discrepancy = sd(bid_discrepancy),
    underestimated = sum(bid_discrepancy > 0),
    overestimated = sum(bid_discrepancy < 0)
  )
```

# 4. Exploratory View

```{r discrepancy-plot}
ggplot(df, aes(x = bid_discrepancy)) +
  geom_histogram(binwidth = 500, fill = "#2C7FB8", color = "white") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#B22222") +
  labs(
    title = "Synthetic Bid-to-Invoice Discrepancy",
    x = "Invoice total minus estimated bill",
    y = "Project count"
  ) +
  theme_minimal()
```

```{r project-type-plot}
ggplot(df, aes(x = reorder(project_type, bid_discrepancy, FUN = median), y = bid_discrepancy)) +
  geom_boxplot(fill = "#74A9CF", color = "#1B1B1B") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#B22222") +
  coord_flip() +
  labs(
    title = "Bid Discrepancy by Synthetic Project Type",
    x = "Project type",
    y = "Bid discrepancy"
  ) +
  theme_minimal()
```

# 5. Model Comparison

```{r model-comparison}
model_full <- lm(
  bid_discrepancy ~ consultant_id + client_id + service_area + project_type +
    department_owner + associate_assigned + billable_travel + billable_shipping +
    candidate_count + estimated_bill + project_cost,
  data = df
)

model_consultant <- lm(bid_discrepancy ~ consultant_id, data = df)
model_client <- lm(bid_discrepancy ~ client_id, data = df)
model_project_type <- lm(bid_discrepancy ~ project_type, data = df)
model_department <- lm(bid_discrepancy ~ department_owner, data = df)
model_operational <- lm(
  bid_discrepancy ~ project_type + department_owner + billable_travel +
    billable_shipping + associate_assigned + candidate_count,
  data = df
)

model_metrics <- function(model, label) {
  pred <- predict(model, df)
  actual <- df$bid_discrepancy
  fstat <- summary(model)$fstatistic
  data.frame(
    model = label,
    adj_r_squared = summary(model)$adj.r.squared,
    mae = mean(abs(actual - pred)),
    rmse = sqrt(mean((actual - pred)^2)),
    p_value = pf(fstat[1], fstat[2], fstat[3], lower.tail = FALSE)
  )
}

bind_rows(
  model_metrics(model_full, "Full diagnostic model"),
  model_metrics(model_consultant, "Consultant group"),
  model_metrics(model_client, "Client group"),
  model_metrics(model_project_type, "Project type"),
  model_metrics(model_department, "Department ownership"),
  model_metrics(model_operational, "Operational model")
) %>%
  mutate(across(where(is.numeric), round, 3))
```

# 6. Interpretation

This workflow is designed for diagnostic review. In a production setting, repeated consultant and client groups would make mixed-effects modeling a better next step than treating all rows as fully independent. The practical use is to support estimate calibration, data-quality improvement, and structured debriefs for high-discrepancy projects.
